---
title: "R Notebook"
output: html_notebook
---
This Notebook performs the analysis of the First part of NUTRISHIEDL STUDY I. Meaning it imports, and analyses all the data from the OBSERVATIONAL part of the study. 



```{r}
#Load Packages
library(ggplot2)
library(readxl)
library(dplyr)
library(reshape2)
library(ggpubr)
library(vtable)
library(network)
library(sna)
library(igraph)
library(ggraph)
library(tidygraph)
library(qgraph)
library(ggraph)
library(mixOmics)
library(grid)

```
First we load all data-sets, and prepare them for analysis
```{r}
#Microbiome data
MicrobiomeData<- read.csv("Statistics/Study1-Data/MicrobiomeData.csv")
MicrobiomeVariables<-colnames(MicrobiomeData)[2:length(colnames(MicrobiomeData))]
#Kidmed scores
KidmedScores <- read.csv("Statistics/Study1-Data/KidmedScores.csv")

#And the conditions
Conditions<-read.csv("Statistics/Study1-Data/Conditions.csv"
)

KidmedScores<-merge(KidmedScores,Conditions,by.x = "Id",by.y = "Id",all = TRUE)
#Merging kidmed and metabolome
MicroKidmed<-merge(KidmedScores,MicrobiomeData,by.x = "Id",by.y = "Id",all = TRUE)

#Metabolome
BileAcids <- read_excel("Statistics/Study1-Data/Metabolome.xlsx",sheet = "BA")
SCFA<-read_excel("Statistics/Study1-Data/Metabolome.xlsx",sheet = "SCFA")

Metabolome<-merge(BileAcids,SCFA,by.x = "Id",by.y = "Id",all = TRUE)
MetabolomeVariables<-colnames(Metabolome)[2:length(colnames(Metabolome))]
ScfaVariables<-colnames(SCFA)[2:length(colnames(SCFA))]
#And now for Food groups
FoodGroups <- read.csv("Statistics/Study1-Data/FoodGroups.csv",
                       stringsAsFactors = FALSE)
FoodGroupsVariables<-colnames(FoodGroups)[2:length(colnames(FoodGroups))]

#We will now merge microbiome and Kidmd with FoodGroup 
FoodMicrobiome<-merge(MicroKidmed,FoodGroups, by.x = "Id",by.y = "Id",all = TRUE)

#And create a complete dataset with all variables
AllVariables<-merge(FoodMicrobiome,Metabolome,by.x = "Id",by.y = "Id",all = TRUE)

#We now load the dataset with the Height, weight and BMI
BmiData <- read_xlsx("Statistics/Study1-Data/BMI.xlsx",sheet = "BMI")
AllVariables<-merge(AllVariables,BmiData,by.x = "Id",by.y = "Id",all = TRUE)

#we now keep only the IDs which were in the Observation study
AllVariables<-AllVariables[which(AllVariables$Id %in% Conditions$Id),]

#We now create a few variables we need
AllVariables$TotalEnergyIntake=
  4*AllVariables$MacroNutrientsCalculation.TotalProteinIntakeG+
  4*AllVariables$MacroNutrientsCalculation.TotalCarbohydrateIntakeG+
  9*AllVariables$MacroNutrientsCalculation.TotalFatIntakeG+
  7*AllVariables$MacroNutrientsCalculation.TotalAlcoholIntakeG
FoodGroupsVariables<-c(FoodGroupsVariables,"TotalEnergyIntake")

NumericalVariables<-colnames(AllVariables)[
  !(colnames(AllVariables)%in%c("Id","Condition","Adherence"))
]
NumericalVariablesData<-AllVariables[NumericalVariables]

VariableTypes<-data.frame(Name=NumericalVariables,Type="None")
VariableTypes$Type[which(VariableTypes$Name %in% colnames(MicrobiomeData))]<-"Microbiome"
VariableTypes$Type[which(VariableTypes$Name %in% colnames(KidmedScores))]<-"Kidmed"
VariableTypes$Type[which(VariableTypes$Name %in% colnames(BileAcids))]<-"BileAcids"
VariableTypes$Type[which(VariableTypes$Name %in% colnames(SCFA))]<-"SCFA"
#VariableTypes$Type[which(VariableTypes$Name %in% colnames(Metabolome))]<-"Metabolome"
VariableTypes$Type[which(VariableTypes$Name %in% colnames(FoodGroups))]<-"FoodGroups"
VariableTypes$Type[which(VariableTypes$Name %in% c("TotalEnergyIntake"))]<-"FoodGroups"

SelectedVariables<-c("Kidmed",MicrobiomeVariables,ScfaVariables,"BMI")

#Last, we parepare a csv with all necessary variables, for sharing where needed
write.csv(x = AllVariables[SelectedVariables,],file = "SupplementaryData1.csv",sep = ","
)
```

#############################################################################################
The following sets of blocks will perform "unidimensional analysis of the data"

In the following block, we prepare the "correltion Matrix" which can be used to plot heatmaps
```{r}
CorrMat<-#na.omit(
  round(cor(AllVariables[,SelectedVariables],
            use="pairwise.complete.obs",method = "spearman"),2)
#)
CorrMat[which(is.na(CorrMat))]<-0
# reorder corr matrix
# using corr coefficient as distance metric
dist <- as.dist((1-CorrMat)/2)
# hierarchical clustering the dist matrix
hc <- hclust(dist,method = "median")
CorrMat <-CorrMat[hc$order, hc$order]

# reduce the size of correlation matrix
melted_cormat <- melt(CorrMat)
melted_cormat$Var1<-as.character(melted_cormat$Var1)
melted_cormat$Var2<-as.character(melted_cormat$Var2)

#We now add a "significance" test to the correlation matrix: 

for(i in 1:dim(melted_cormat)[1]){
  Pval<-cor.test(AllVariables[,melted_cormat$Var1[i]],
                 AllVariables[,melted_cormat$Var2[i]],method="spearman")$p.value
  if(!is.numeric(Pval) || is.na(Pval) || Pval > 0.05){
    Significance<-NA
  }else if(Pval > 0.01){
    Significance<-1
  }else if(Pval > 0.001){
    Significance<-2
  }else if(Pval <= 0.001){
    Significance<-3
  }else{
    Significance<-NA
  }
  
  melted_cormat$Pvalue[i]<-Pval
  melted_cormat$Significance[i]<-Significance
}

melted_cormat$Significance<-as.numeric(melted_cormat$Significance)

```

We now create Boxplots for All variables in relationship to the conditions. These are Figures 2, 3 and 4. 
```{r}
#my_comparisons <- list( c("HC", "Obese"), c("HC", "T1D"), c("T1D", "Obese") )
#Micro<-colnames(MicrobiomeData)[9]
#my_comparisons<-list( c("HC", "Obese"), c("Obese", "T1D"), c("HC", "T1D") )
#for (Variable in colnames(AllVariables)[!(colnames(AllVariables)%in%c("Id","Condition","Adherence"))]){
Plots<-list()
for (Variable in SelectedVariables){
  plot<-ggplot(data = AllVariables,aes(x=Condition,y=get(Variable)))+
    geom_boxplot(notch = TRUE,aes(fill=Condition))+
    theme_bw()+
    #stat_compare_means(comparisons = my_comparisons, method = "kruskal.test")
    stat_pwc(label = "p.format", method = "wilcox.test",
             hide.ns = TRUE,p.adjust.method = "none",size = 1.5,label.size = 7)+
    theme(
      legend.position = "none",
      text=element_text(size=20),
      axis.title.x = element_blank(),
      axis.text = element_text(size=20),
    )+
    scale_fill_manual(values = c("steelblue","tomato","plum"))
  if(Variable %in% MetabolomeVariables){
    plot<-plot+ylab(paste(Variable,"[µmol/g creatinine]"))
  }else if(Variable %in% MicrobiomeVariables){
    plot<-plot+ylab(paste(Variable,"[relative abundance]"))
  }else{
    plot<-plot+ylab(Variable)
  }
  
  #We now save the plots in the different folders according to data type. 
  
  #The microbiome BoxPlots are the ones shown in Figure 3
  if(Variable %in% MicrobiomeVariables){
    ggsave(filename = paste("Statistics/Plots/BoxPlotsConditionsMicrobiome/",Variable,".pdf",sep = ""),
           plot = plot)
    
    
  }else if(Variable %in% ScfaVariables){
    #Next, we store the Scfa data which is shown in Figure 4
    ggsave(filename = paste("Statistics/Plots/BoxPlotsConditionsScfa/",Variable,".pdf",sep = ""),
           plot = plot)
    
    
  }else{ 
    #The remaining variables include the KIDMED, which is shown in Figure 2
    ggsave(filename = paste("Statistics/Plots/BoxPlotsConditionsSelect/",Variable,".pdf",sep = ""),
           plot = plot)
  }
}

```

We can now make  series of "heatmaps" where we plot several variables and color code the different correlations. This is Figure 8
```{r}
#Plot of all variables
MicrobiomeVariablesPlot<-MicrobiomeVariables[which(MicrobiomeVariables != "animalis.lactis")]
TilePlot<-ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value))+ 
  geom_tile()+
  theme_bw() + #xlab("Varible 1") + ylab("Variable 2")+ 
  xlab("") + ylab("")+
  coord_equal(ratio = 1)+
  scale_x_discrete(expand = c(0,0),limits = c(ScfaVariables,MicrobiomeVariablesPlot))+
  scale_y_discrete(expand = c(0,0),limits = c(ScfaVariables,MicrobiomeVariablesPlot)) +
  scale_fill_gradient2(low = "blue", mid="white", high = "red",
                       limits=c(-1,1),name="Correlation")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 10))+
  ggtitle("Correlation between variables")+
  geom_point(aes(size=Significance),fill="black")+
  scale_size_continuous(limits = c(0,3),
                        range=c(0,2),
                        name="Significance",
                        labels=c("Not significant","P-Value < 5%","P-value < 1%",
                                 "P-value < 0.1%"))+
  # geom_rect(aes(xmin="Acetic.acid",ymin="Acetic.acid",xmax="Isoleucine",ymax="Isoleucine"),color="black",size=1)
  geom_rect(aes(xmin=0.5,ymin=0.5,xmax=12.5,ymax=12.5),color="black",size=0.7,fill=NA)+
  geom_rect(aes(xmin=12.5,ymin=12.5,xmax=19.5,ymax=19.5),color="black",size=0.7,fill=NA)
# ggsave(filename = "Statistics/TilePlot.pdf",plot = TilePlot,units = "px",width = 5000,height = 5000
# )
TilePlot
```

We will now follow with Partial Least Square Discriminant analysis (PLS-DA), to find the predictors for the different conditions. The first one (Microbiome), correspods to Figure 5

```{r}
#Begin by selecting the independent and the response variables
SelectedColors<-c("steelblue","tomato","plum")
FittedVariables<-AllVariables[,MicrobiomeVariables]
Response<-AllVariables[,"Condition"]
ModelPLS<-splsda(FittedVariables,Response,scale = TRUE,ncomp = 2)

background.mahal <-  background.predict(ModelPLS,
                                        comp.predicted = 2,
                                        dist = 'mahalanobis.dist')
plotIndiv(ModelPLS, 
          rep.space = 'XY-variate', 
          ellipse = TRUE,  # plot using the ellipses
          legend = TRUE,
          background = background.mahal,
          ind.names=FALSE,
          title = "PLS-DA Microbiome Variables \n Scatter plot",
          size.title = rel(1),
          X.label = "LV 1",
          Y.label = "LV 2",
          style = "ggplot2",
          col.per.group = SelectedColors
)

# plotIndiv(ModelPLS, 
#           rep.space = 'multi', 
#           # ellipse = TRUE,  # plot using the ellipses
#           legend = TRUE,
#           background = background.mahal)
PlotLv1<-plotLoadings(ModelPLS, method = 'mean', contrib = 'max',comp = 1,
                      title="Individual variable contributions on LV 1",
                      size.title = rel(1),legend.color = SelectedColors
)$X


# PlotLv1$Var<-c(
#   "B. breve",
#   "B. bifidum",
#   "B. longum subs. infantis",
#   "B. longum subs. longum",
#   "B. animalis subs. animalis",
#   "B. pseudocatenulatum",
#   "B. adolescentis"    
# )
# 
# Lv1<-ggplot(data = PlotLv1,aes(x=importance,y=Var,fill=GroupContrib))+
#   geom_col()+
#   scale_fill_manual(values = SelectedColors,name="Outcome")+
#   theme_bw()+
#   theme(
#     axis.title.y = element_blank(),
#   )
# Lv1  

PlotLv2<-plotLoadings(ModelPLS, method = 'mean', contrib = 'max',comp = 2,
                      title="Individual variable contributions on LV 2",
                      size.title = rel(1),legend.color = SelectedColors
)

```
Following the results above, we see that the LV1 associated with microbiome, is explaining Obesity quite well. As a validation of the model, we propose the following strategy: 
We will compute the value of LV1 for each participant, and correlate with his/her BMI. 
This is shown in Figure 6

```{r}
#We start by computing LV1 in each participant we do this by creating 2 matrixes, and then using the vector multiplication operation
A<-as.matrix(AllVariables[,MicrobiomeVariables])
B<-matrix(ModelPLS[["loadings.star"]][[1]][,1])
AllVariables$LV1<-A%*%B
x<-"BMI"
y<-"LV1"

ScatterPlot<-ggplot(data = AllVariables,
                    aes(x=get(x),y=get(y)))+
  geom_point(size=3,aes_string(colour=ColorVariable))+
  geom_smooth(method=lm,se=FALSE)+
  theme_bw()+xlab(x)+ylab(y)+
  ggtitle(paste(x,"VS",y,"\n","Cor=",
                round(cor(AllVariables[,x],AllVariables[,y],
                          use = "pairwise.complete.obs",method = "spearman"
                ),digits = 2),"P-value=",
                round(cor.test(AllVariables[,x],AllVariables[,y],method="spearman"
                               
                )$p.value,digits = 3)
  ))+
  scale_color_manual(values = c("steelblue","tomato","plum"))+
  theme(
    #      legend.position = "none",
    text=element_text(size=20),
    #      axis.title.x = element_blank(),
    axis.text = element_text(size=20)
  )
ggsave(filename = 
         paste("Statistics/Plots/",
               x,"-",y,".pdf",sep = ""),plot = ScatterPlot)
ScatterPlot

```


We now move to do the same PLS-DA analysis for metabolic data, which is shown in Figure 7

```{r}
#Begin by selecting the independent and the response variables
KeyMetabolites<-c("Acetic.Acid",
                  "Propionic.Acid",
                  "Isobutyric.Acid",
                  "Butyric.Acid",
                  "Methylbutyric.Acid",
                  "Isovaleric.Acid",
                  "Valeric.Acid",
                  "Heptanoic.Acid",
                  "Valine",
                  "Leucine",
                  "Isoleucine")
FittedVariables<-AllVariables[,c(ScfaVariables)]
Response<-AllVariables[,"Condition"]
ModelPLS<-splsda(FittedVariables,Response,scale = TRUE,ncomp = 2)

background.mahal <-  background.predict(ModelPLS,
                                        comp.predicted = 2,
                                        dist = 'mahalanobis.dist')


plotIndiv(ModelPLS, 
          rep.space = 'XY-variate', 
          ellipse = TRUE,  # plot using the ellipses
          legend = TRUE,
          background = background.mahal,
          ind.names=FALSE,
          title = "PLS-DA Metabolic variables \n Scatter plot",
          size.title = rel(1),
          X.label = "LV 1",
          Y.label = "LV 2",
          col.per.group = SelectedColors
)

# 
# plotIndiv(ModelPLS, 
#           rep.space = 'multi', 
#           # ellipse = TRUE,  # plot using the ellipses
#           legend = TRUE,
#           background = background.mahal)
plotLoadings(ModelPLS, method = 'mean', contrib = 'max',comp = 1,
             title="Individual variable contributions on LV 1",
             size.title = rel(1),legend.color = SelectedColors)
plotLoadings(ModelPLS, method = 'mean', contrib = 'max',comp = 2,
             title="Individual variable contributions on LV 2",
             size.title = rel(1),legend.color = SelectedColors)

```


